Parameter Inference\\with Bifurcation Diagrams

### 2021-10-23 docs/appendix.tex
overleaf

\section{Properties of the Bifurcation Measure}
\label{appendix:measure}

@@

% \section{Simplification of two-state model}

% Consider the general two-state model:
% \begin{equation}
%     \dfrac{dv_i}{dt} = \dfrac{a_i + b_i (K_i v_j)^2}{1 + (K_i v_j)^2} - \mu_i v_i
% \end{equation}

% Without loss of generality, we can rescale $v_i = U_i u_i$. Then,
% \begin{equation}
%     \dfrac{U_idu_i}{dt} = \dfrac{a_i + b_i (K_i U_j u_j)^2}{1 + (K_i U_j u_j)^2} - \mu_i U_i u_i
% \end{equation}

% Dividing through by $U_i$, we obtain
% \begin{equation}
%     \dfrac{du_i}{dt} = \dfrac{\frac{a_i}{U_i} + \frac{b_i}{U_i} (K_i U_j u_j)^2}{1 + (K_i U_j u_j)^2} - \mu_i u_i
% \end{equation}

% By choosing $U_i = b_i$, the model reduces to:
% \begin{equation}
%     \dfrac{du_1}{dt} = \dfrac{\hat{a}_1 + (p u_2)^2}{1 + (p u_2)^2} - \mu_1 u_1, \quad
%     \dfrac{du_2}{dt} = \dfrac{\hat{a}_2 + (k u_1)^2}{1 + (k u_1)^2} - \mu_2 u_2
% \end{equation}
% where $p = \frac{K_1}{\sqrt{b_1}}$ and $k = \frac{K_2}{\sqrt{b_2}}$.

>>>>

\clearpage
\section{Bifurcation Measure Properties}
\label{appendix:conditions}
Consider a vector $v(s)\in\Reals^N$ parameterised by $s\in\Reals$ that is tangent to an equilibrium manifold defined by $\rates(u)=0$. The conditions for a non-degenerate static bifurcation at $s^*$ along such a tangent can be expressed in terms of an eigenvalue $\lambda(s)$ of the state-space Jacobian crossing zero with a finite slope. A bifurcation exists at $s^*$ if
\begin{align}
    \frac{\partial\rates}{\partial u}v(s)=\lambda(s) \,v(s)
    \quad\exists\lambda:\quad
    \left.\lambda(s)\right|_{s=s^*}=0
    \qquad
    \left.\frac{d\lambda}{ds}\right|_{s=s^*}\neq 0
    \label{eq:conditions}
\end{align}
These conditions are necessary and sufficient for a non-degenerate static local breakdown of stability. For now we do not consider dynamic bifurcations involving limit cycles or imaginary parts of eigenvalues and restrict $\lambda\in\Reals$. Cases where both $\left.\lambda(s)\right|_{s=s^*}=0$ and $\left.\frac{d\lambda}{ds}\right|_{s=s^*}=0$ require investigation into higher order derivatives $\frac{d^n\lambda}{ds^n}$. These are the cases we refer to as \emph{degenerate} and are not considered here.

Instead of considering conditions on each eigenvalue individually it is possible to use the determinant of the state-space Jacobian to detect whether the conditions \eqref{eq:conditions} are satisfied. The determinant can be expressed as the product of eigenvalues
\begin{equation}
    \Det=\prod_{n=1}^N\lambda_n(s)
    \label{eq:determinant}
\end{equation}
Applying the product rule when differentiating yields
\begin{align}
    \frac{d}{ds}\Det&=
    \sum_{n=1}^N\frac{d\lambda_n}{ds}\prod_{n'\neq n}\lambda_{n'}(s)\\
    &=\Det\sum_{n=1}^N\frac{d\lambda_n}{ds}\lambda_{n}(s)^{-1}
\end{align}
Substituting this expression into measure \eqref{eq:measure}
\begin{equation}
    \measure(s)=
    \left(1+\left|\sum_{n=1}^N\frac{d\lambda_n}{ds}\lambda_{n}(s)^{-1}\right|^{-1}\right)^{-1}
\end{equation}
Which implies the following
\begin{equation}
    \exists\lambda:\quad
    \begin{cases}
        \,\lambda(s)=0 \quad\frac{d\lambda}{ds}\neq 0\\
        \,\lambda(s)\neq0 \quad\frac{d\lambda}{ds}\rightarrow\pm\infty
    \end{cases}
    \implies
    \measure(s)=1
\end{equation}
If there exists an eigenvalue that satisfies conditions \eqref{eq:conditions} then the measure is equal to one. The measure also approaches one in cases where the rate of change of an eigenvalue with respect to a manifold $s$ location  diverges while not crossing zero. This gives rise to finite gradients in the eigenvalue term in regimes far away from any bifurcation.

\section{Extension for Hopf Bifurcations}
\label{appendix:hopf}

In order to detect bifurcations involving limit cycles, the measure must be extended to detect changes in the real part $\Re\mathrm{e}[\lambda(s)]$ for any eigenvalue of the Jacobian. These conditions can no longer be compactly written in terms of the determinant. Instead, the measure can be defined as the sum of eigenvalue terms 
\begin{align}
    \measure(s):=\sum_{\lambda(s)\in\frac{\partial\rates}{\partial u}}
    \left(\left|\frac{d}{ds}\log\Re\mathrm{e}[\lambda(s)]\right|^{-1}+1\right)^{-1}
    \label{eq:hopf-measure}
\end{align}

The directional derivative of the logarithm diverges under two conditions: when eigenvalues vanish $\lambda(s)=0$ and when the directional derivative $\frac{d}{ds}\Re\mathrm{e}[\lambda(s)]$ diverges. These properties are sufficient for detecting the onset of damped oscillations and emergence of limit cycles via Hopf bifurcation as shown in Figure \ref{fig:hopf-measure}. Eigenvalues with negative real part which gain a finite imaginary part give rise to damped oscillations. At this onset we observe a discontinuity in the derivative $\frac{d}{ds}\Re\mathrm{e}[\lambda(s)]$ which is detected by equation \eqref{eq:hopf-measure}. Once damped oscillations exist, flipping the stability of the stable fixed point gives rise to a limit cycle, which can be detected by inspecting $\Re\mathrm{e}[\lambda(s)]$.

\begin{figure}[H]
\centering
    \includegraphics[width=60mm]{hopf-measure}
    \caption{Bifurcation measure $\measure(s)$ and eigenvalues $\lambda(s)$ along the arclength $s$ for two different bifurcation curves demonstrating how the measure detects non-zero imaginary parts $\Im\mathrm{m}[\lambda]$ (onset of damped oscillations marked by circle) and sign changes in real parts $\Re\mathrm{e}[\lambda]$ (Hopf bifurcations marked by stars)}
    \label{fig:hopf-measure}
\end{figure}

In principle it is possible to construct measures to detect a variety of bifurcations as long as the conditions can be expressed in terms of derivatives with respect to fixed-point manifold direction $s$. Measures can be used sequentially or in parallel to encourage optimisers to run through a sequence of bifurcations or place specific bifurcation types next to each other.

@@

\section{A More Complex Model}
\label{appendix:double-exclusive}
To demonstrate the wider reaching applicability of our method we optimise the \emph{double exclusive reporter} \cite{Grant2020InterpretationCircuit}, a synthetic \emph{E. Coli} designed to exhibit a cusp bifurcation with respect to two input signals. Figure \ref{fig:double-exclusive-trajectories} shows the cost function with respect to iteration for sample trajectories. We note that due to a

\section{Steady State Solvers}
\label{appendix:stedy-state-solvers}

### 2021-10-23 docs/article.tex
overleaf

% \usepackage{neurips_2021}
\usepackage[preprint]{neurips_2021}
% \usepackage[final]{neurips_2021}

@@

\title{Parameter Inference\\with Bifurcation Diagrams}
\author{
        Gregory Szep\\King's College London\\London, WC2R 2LS\\
        \texttt{gregory.szep@kcl.ac.uk}
    \And
        Attila Csik\'asz-Nagy\\King's College London\\London, WC2R 2LS\\
        \texttt{attila.csikasz-nagy@kcl.ac.uk}
    \And
        Neil Dalchau\\Microsoft Research Cambridge\\Cambridge, CB1 2FB\\
        \texttt{ndalchau@microsoft.com}
}

@@

\maketitle
\begin{abstract}
    Estimation of parameters in differential equation models can be achieved by applying learning algorithms to quantitative time-series data. However, sometimes it is only possible to measure qualitative changes of a system in response to a controlled condition. In dynamical systems theory, such change points are known as \textit{bifurcations} and lie on a function of the controlled condition called the \textit{bifurcation diagram}. In this work, we propose a gradient-based semi-supervised approach for inferring the parameters of differential equations that produce a user-specified bifurcation diagram. The cost function contains a supervised error term that is minimal when the model bifurcations match the specified targets and an unsupervised bifurcation measure which has gradients that push optimisers towards bifurcating parameter regimes. The gradients can be computed without the need to differentiate through the operations of the solver that was used to compute the diagram. We demonstrate parameter inference with minimal models which explore the space of saddle-node and pitchfork diagrams and the genetic toggle switch from synthetic biology. Furthermore, the cost landscape allows us to organise models in terms of topological and geometric equivalence.
\end{abstract}

@@

Inferring the parameters of a model directly from a bifurcation diagram is difficult because it is not obvious how multiple  parameters in concert control the existence and position of a bifurcation. It could even be impossible for the model to bifurcate in the manner desired. Several approaches exist to place bifurcations to desired locations once a manifold is present \cite{Iwasaki1997AnType,Lu2006InverseSystems,Dobson2004DistanceBifurcations} yet typically resort to sampling techniques to search for them in the first place \cite{Chickarmane2005BifurcationTool,Conrad2006BifurcationClock}. Progress has been made in cases where model structure and stability conditions are used to refine the search space \cite{Otero-Muras2018Optimization-basedModels,Otero-Muras2014ACurves} yet the resulting objectives are still not explicit in the bifurcation targets and also not differentiable. In the emerging field of scientific machine learning \cite{Rackauckas2017Differentialequations.jlJulia,Rackauckas2018ASolutions,Rackauckas2020UniversalLearning}, parameters of structured mechanistic models are favoured over flexible models in larger parameter spaces. A scalable method for navigating the space of bifurcation diagrams would enable design of differential equations with high-level qualitative constraints. Furthermore one could begin organising models according to qualitatively distinct behaviours.

@@

The problem of inferring differential equation parameters against a user-specified bifurcation diagram decomposes into two parts: searching for bifurcating regimes and matching the locations of bifurcation points to desired values. Matching bifurcation locations is a supervised problem where the data are expressed as bifurcations points \cite{Lu2006InverseSystems,Conrad2006BifurcationClock}. Searching for bifurcations is an unsupervised problem because when bifurcations are not present, there is no distance defined between data and prediction \cite{Chickarmane2005BifurcationTool}. Therefore only properties of the model can be used to start the search. We propose a semi-supervised approach for performing both tasks in an end-to-end fashion. The bifurcation diagram encodes high-level qualitative information defined by state space structures, rather than kinetics. We apply the strategy of implicit layers \cite{Look2020DifferentiableLayers,Bai2019DeepModels} to calculate gradients. To compute the diagram we use a predictor-corrector method called deflated continuation \cite{Farrell2016TheDiagrams,Veltz2020BifurcationKit.jl} developed for partial differential equations.

@@

For the purposes of introducing this work, we will consider the simplest class of bifurcations known as \textit{co-dimension one} bifurcations not including limit cycles. Therefore $\targets$ should contain conditions for which we hypothesise changes in multi-stable behaviour. Let the equations be
\begin{align}
	\frac{\partial u}{\partial t}=\rates(u,p)
	\qquad\mathrm{where}\quad
	\rates : \Reals^{N+1}\rightarrow\Reals^N
	\label{eq:model}
\end{align}
In the context of the differential equations, and not considering limit cycles for now, a co-dimension one bifurcation can be defined by a set of conditions on the determinant of the Jacobian $\Det$. The determinant of the Jacobian quantifies the rate at which trajectories in a local patch of state-space $u\in\Reals^N$ converge or diverge. Let $s\in\Reals$ parametrise the curves that trace out the bifurcation diagram. Any location on the curve $u(s)$ and $p(s)$ must satisfy the steady-state of equations \eqref{eq:model}. The determinant approaching zero along the diagram means that the dynamics of the system are slowing down, which is an important indicator for the onset of a transition between qualitative behaviours. Furthermore, the slowing down must necessarily be followed by a breakdown of stability; for this to be true it is sufficient \textit{but not necessary} to require that the determinant cross zero with a finite slope, meaning that its directional derivative along the diagram $\frac{d}{ds}\Det$ is not zero. The set of predicted values for the control condition $\predictions(\theta)\subset\Reals$ at which bifurcations occur are defined as
\begin{align}
	\predictions(\theta):=\left\{\,
	p\,\,\exists\,\,u:\,\,\rates(u,p)=0,\,\,\Det=0,
	\,\, \frac{d}{ds}\Det\neq0
	\,\right\}
	\label{eq:predictions}
\end{align}
A proof of how the conditions \eqref{eq:predictions} are sufficient but not necessary for local bifurcations is detailed in Appendix \ref{appendix:conditions}. The most common bifurcations between steady states, not including limit cycles, are saddle-nodes and pitchforks \cite{Haragus2011LocalSystems}. Saddle-node bifurcations, which often appear in pairs (Figure \ref{fig:saddle-node}) are defined by stable and unstable fixed points meeting and disappearing. Pitchfork bifurcations are special cases of the saddle-node where a single steady state splits into two stable and one unstable steady state (Figure \ref{fig:pitchfork}). To illustrate these bifurcations, we define minimal models (Figure \ref{fig:minimal-models}) that span the space of saddle-node and pitchforks, where indeed zero crossings in the determinant with a finite slope define the set of prediction $\predictions(\theta)$. The location of these crossings in general may not match the targets $\targets$.

@@

\section{Proposed Method}
\label{section:method}
\subsection{Semi-supervised Cost Function}

To identify parameter sets that give rise to bifurcation diagrams with specified bifurcation points, we propose a semi-supervised cost function that comprises two terms. The role of the supervised term is simply to reward predicted bifurcations to coincide with the specified target locations. This of course relies on such bifurcations existing. The role of the unsupervised term is to encourage an optimiser to move towards parameter regimes that do exhibit bifurcations.

\subsubsection{Supervised term: matching bifurcations to target locations}

In order for predicted bifurcations $p(\theta)\in\predictions(\theta)$ to match targets $p'\in\targets$ we need to evaluate an error term $|p(\theta)-p'|$. A naive approach might take an average over the norms for all prediction-target pairs. However this gives rise to unwanted cross-terms and the possibility of multiple predictions matching the same target without any penalty for unmatched targets. Therefore, we choose a geometric mean over the predictions and an arithmetic mean over targets:
\begin{equation}
    \error(\theta, \targets) = \frac{1}{|\targets|}\sum_{p'\in\targets}
    \prod_{p(\theta)\in\predictions(\theta)}|p(\theta)-p'|
    ^{\frac{1}{|\predictions|}}
    \label{eq:error}
\end{equation}
The supervised term is only zero when each target is matched by at least one prediction and allows for cases where the number of predictions is greater than or equal to the number of targets $|\predictions| \geq |\targets|$. An alternative approach, which undesirably introduces more hyper-parameters, would be to let each prediction $\predictions(\theta)$ represent the centroid of a mixture distribution and use expectation-maximisation to match the centroids to targets $\targets$.

\subsubsection{Unsupervised term: encouraging bifurcations}
\begin{wrapfigure}{r}{0.5\textwidth}
    \centering
    \includegraphics[width=\linewidth]{bifurcation-measure}
    \caption{Bifurcation measure $\measure(s)$ and determinant $\Det$ along the arclength $s$ of two different bifurcation curves demonstrating how maximising the measure along the curve maintains the existing bifurcation marked by a circle, while encouraging new bifurcations marked by stars.}
    \label{fig:measure}
\end{wrapfigure}

@@

% Finally, we note that $\measure(s)$ has the desirable property of being invariant to certain classes of transformations on $\rates(u,p)$ that do not change bifurcation locations (details provided in Appendix \ref{appendix:measure}).

@@

Equipped with a measure that quantifies the appearance of bifurcations along a bifurcation arc we can define the total measure for a bifurcation diagram as
\begin{equation}
    \Psi(\theta):=\frac{
        \int_{\rates(u,p)=0}\!
        \measure(s)\,\mathrm{d}s
    }{
        \int_{\rates(u,p)=0}\!\
        \mathrm{d}s
    }.
    \label{eq:total-measure}
\end{equation}
Here we denote $\int_{\rates(u,p)=0}\mathrm{d}s$ as the sum of the line integrals in $(u,p)\in\Reals^{N+1}$ defined by the level set $\rates(u,p)=0$ with $s$ being an arbitrary parametrisation of the curves. The total measure $\Psi(\theta)$ is normalised such that $\Psi(\theta)\rightarrow1$ in the regimes where the controlled condition region $p$ is densely packed with bifurcations. The total measure $\Psi(\theta)$ is added to the supervised term as if it were a likelihood. This defines the semi-supervised cost function as
\begin{align}
    \loss(\theta|\targets):=
    \big(|\predictions|-|\targets|\big)\log\Psi(\theta)+
    \error(\theta, \targets),
    \label{eq:loss}
\end{align}
The pre-factor $|\targets|-|\predictions|$ in the unsupervised term ensures that the gradients are always pushing optimisers towards a state where $|\targets|=|\predictions|$. This can be seen as a step-wise annealing of the unsupervised term until the desired state is reached.

\subsection{Differentiating the semi-supervised cost function}

To make use of gradient-based optimisers to locate desired bifurcation diagrams, we show here how to differentiate the cost function. First, we note that while individual bifurcations $p(\theta)$ depend smoothly on $\theta$, the total number of predictions $|\predictions|$ does not have gradient contributions with respect to $\theta$. Therefore, we can safely drop the dependency in the prediction counter and now proceed in taking gradients with respect to $\theta$ knowing that the only dependencies we need to track are for individual bifurcations $p(\theta)$ within the definition the supervised term \eqref{eq:error} and the total measure \eqref{eq:total-measure}. Therefore, 
\begin{align}
    \frac{\partial\loss}{\partial\theta}=
    \big(|\predictions|-|\targets|\big)\,\lambda\,
    \frac{\partial \Psi}{\partial\theta}\Psi(\theta)^{-1}+
    \frac{1}{|\targets||\predictions|}\sum_{p'}
    \prod_{p(\theta)}|p(\theta)-p'|^{\frac{1}{|\predictions|}}
    \sum_{p(\theta)}\frac{\partial p}{\partial\theta}\left(p(\theta)-p'\right)^{-1}
    \label{eq:gradient-loss}
\end{align}
In a similar vein to back-propagation through neural differential equations \cite{Chen2018NeuralEquations} we would like to be able to calculate the gradient $\frac{\partial\loss}{\partial\theta}$ without having to differentiate through the operations of the solver that finds the bifurcation diagram $\rates(u,p)=0$ and the bifurcation locations $p(\theta)$. To calculate the gradient of the measure $\frac{\partial \Psi}{\partial\theta}$ we need to differentiate line integrals that depend on $\theta$. Fortunately this can be done by the application of the generalised Leibniz integral rule, details of which can be found in Appendix \ref{appendix:leibniz-rule}.

@@

@@

We proposed a gradient-based semi-supervised approach for inferring the parameters of differential equations that produce a user-specified bifurcation diagram. By applying implicit layers \cite{Look2020DifferentiableLayers,Bai2019DeepModels} and the generalised Leibniz rule \cite{Flanders1973DifferentiationSign} to the geometry of the implicitly defined steady states \cite{Goldman2005CurvatureSurfaces} it is possible to use automatic differentiation methods to efficiently calculate gradients. We defined a bifurcation measure that uses the determinant of the state-space Jacobian as an indicator for bifurcating parameter regimes in the unsupervised term of the cost function. The gradients of the cost can be efficiently computed using automatic differentiation methods. The computational bottleneck is the evaluation of the state-space Jacobian determinant which limits the implementation to ordinary differential equations.

@@

Although we did not consider limit cycles, our approach can be modified using the conditions for Poincar\'e-Andronov-Hopf bifurcations instead of the conditions on the Jacobian determinant. Our approach generalises naturally to bifurcation manifolds such as limit point curves or surfaces. This is because the normal components of implicit derivatives can still be calculated for under-determined systems of equations \cite{Jos2011OnSurface,Tao2016Near-IsometricTracking,Fujisawa2013CalculationInvariance}. In the case of manifolds it would be more appropriate to use isosurface extraction algorithms rather than continuation to obtain the steady-state manifold. Our approach does not depend on the details of the steady-state solver and therefore can still be applied.

>>>>

% \usepackage{neurips_2021}
% \usepackage[preprint]{neurips_2021}
\usepackage[final]{neurips_2021}

@@

\title{Parameter Inference\\with Bifurcation Diagrams}
\author{
        Gregory Szep\\King's College London\\London, WC2R 2LS, UK\\
        \texttt{gregory.szep@kcl.ac.uk}
    \And
        Attila Csik\'asz-Nagy\\P\'azm\'any P\'eter Catholic University\\Budapest, 1083, Hungary\\
        \texttt{csikasznagy@gmail.com}
    \And
        Neil Dalchau\\Microsoft Research Cambridge\\Cambridge, CB1 2FB, UK\\
        \texttt{ndalchau@gmail.com}
}

@@

\maketitle
\begin{abstract}
    Estimation of parameters in differential equation models can be achieved by applying learning algorithms to quantitative time-series data. However, sometimes it is only possible to measure qualitative changes of a system in response to a controlled condition. In dynamical systems theory, such change points are known as \textit{bifurcations} and lie on a function of the controlled condition called the \textit{bifurcation diagram}. In this work, we propose a gradient-based approach for inferring the parameters of differential equations that produce a user-specified bifurcation diagram. The cost function contains an error term that is minimal when the model bifurcations match the specified targets and a bifurcation measure which has gradients that push optimisers towards bifurcating parameter regimes. The gradients can be computed without the need to differentiate through the operations of the solver that was used to compute the diagram. We demonstrate parameter inference with minimal models which explore the space of saddle-node and pitchfork diagrams and the genetic toggle switch from synthetic biology. Furthermore, the cost landscape allows us to organise models in terms of topological and geometric equivalence.
\end{abstract}

@@

Inferring the parameters of a model directly from a bifurcation diagram is difficult because it is not obvious how multiple  parameters in concert control the existence and position of a bifurcation. It could even be impossible for the model to bifurcate in the manner desired. Several approaches exist to place bifurcations to desired locations once a manifold is present \cite{Iwasaki1997AnType,Lu2006InverseSystems,Dobson2004DistanceBifurcations} yet typically resort to sampling techniques to search for them in the first place \cite{Chickarmane2005BifurcationTool,Conrad2006BifurcationClock}. Progress has been made in cases where model structure and stability conditions are used to refine the search space \cite{Otero-Muras2018Optimization-basedModels,Otero-Muras2014ACurves} yet the resulting objectives are still not explicit in the bifurcation targets and also not differentiable. It is always possible to bias goodness of fit during optimisation with domain-specific knowledge \cite{}, however these approaches do not aim to generalise across a wider set of qualitative behaviours. In the emerging field of scientific machine learning \cite{Rackauckas2017Differentialequations.jlJulia,Rackauckas2018ASolutions,Rackauckas2020UniversalLearning}, parameters of structured mechanistic models are favoured over flexible models in larger parameter spaces. A scalable method for navigating the space of bifurcation diagrams would enable design of differential equations with high-level qualitative constraints. Furthermore one could begin organising models according to qualitatively distinct behaviours.

@@

The problem of inferring differential equation parameters against a user-specified bifurcation diagram decomposes into two parts: searching for bifurcating regimes and matching the locations of bifurcation points to desired values. Matching bifurcation locations is a supervised problem where the data are expressed as bifurcations points \cite{Lu2006InverseSystems,Conrad2006BifurcationClock}. Searching for bifurcations is an unsupervised problem because when bifurcations are not present, there is no distance defined between data and prediction \cite{Chickarmane2005BifurcationTool}. Therefore only properties of the model can be used to start the search. We propose an approach for performing both tasks in an end-to-end fashion. The bifurcation diagram encodes high-level qualitative information defined by state space structures, rather than kinetics. We apply the strategy of implicit layers \cite{Look2020DifferentiableLayers,Bai2019DeepModels} to calculate gradients. To compute the diagram we use a predictor-corrector method called deflated continuation \cite{Farrell2016TheDiagrams,Veltz2020BifurcationKit.jl} developed for partial differential equations.

@@

For the purposes of introducing this work, we will consider the simplest class of bifurcations known as \textit{co-dimension one} bifurcations not including limit cycles. Therefore $\targets$ should contain conditions for which we hypothesise changes in multi-stable behaviour. Let the equations be
\begin{align}
	\frac{\partial u}{\partial t}=\rates(u,p)
	\qquad\mathrm{where}\quad
	\rates : \Reals^{N+1}\rightarrow\Reals^N
	\label{eq:model}
\end{align}
In the context of the differential equations, and not considering limit cycles for now, we show that a static non-degenerate bifurcation can be defined by a set of conditions on the determinant of the Jacobian $\Det$. The determinant of the Jacobian quantifies the rate at which trajectories in a local patch of state-space $u\in\Reals^N$ converge or diverge. Let $s\in\Reals$ parametrise the curves that trace out the bifurcation diagram. Any location on the curve $u(s)$ and $p(s)$ must satisfy the steady-state of equations \eqref{eq:model}. The determinant approaching zero along the diagram means that the dynamics of the system are slowing down, which is an important indicator for the onset of a transition between qualitative behaviours. Furthermore, the slowing down must necessarily be followed by a breakdown of stability; for this to be true it is sufficient \textit{but not necessary} to require that the determinant cross zero with a finite slope, meaning that its directional derivative along the diagram $\frac{d}{ds}\Det$ is not zero. This is the non-degeneracy condition. The set of predicted values for the control condition $\predictions(\theta)\subset\Reals$ at which bifurcations occur are defined as
\begin{align}
	\predictions(\theta):=\left\{\,
	p\,\,\exists\,\,u:\,\,\rates(u,p)=0,\,\,\Det=0,
	\,\, \frac{d}{ds}\Det\neq0
	\,\right\}
	\label{eq:predictions}
\end{align}
A proof of how the conditions \eqref{eq:predictions} are necessary and sufficient for static non-degenerate bifurcations is detailed in Appendix \ref{appendix:conditions}. The most common bifurcations between steady states, not including limit cycles, are saddle-nodes and pitchforks \cite{Haragus2011LocalSystems}. Saddle-node bifurcations, which often appear in pairs (Figure \ref{fig:saddle-node}) are defined by stable and unstable fixed points meeting and disappearing. Pitchfork bifurcations occur where a single steady state splits into two stable and one unstable steady state (Figure \ref{fig:pitchfork} shows an \textit{imperfect} pitchfork; a \textit{perfect} pitchfork arises when $\theta_1=0$). To illustrate these bifurcations, we define minimal models (Figure \ref{fig:minimal-models}) that span the space of saddle-node and pitchforks, where indeed zero crossings in the determinant with a finite slope define the set of prediction $\predictions(\theta)$. The location of these crossings in general may not match the targets $\targets$.

@@

\section{Proposed Method}
\label{section:method}
\subsection{Cost Function}

To identify parameter sets that give rise to bifurcation diagrams with specified bifurcation points, we propose a cost function that comprises two terms. The role of the error term is simply to reward predicted bifurcations to coincide with the specified target locations. This of course relies on such bifurcations existing. The role of the eigenvalue term is to encourage an optimiser to move towards parameter regimes that do exhibit bifurcations.

\subsubsection{Error term: matching bifurcations to target locations}

In order for predicted bifurcations $p(\theta)\in\predictions(\theta)$ to match targets $p'\in\targets$ we need to evaluate an error term $|p(\theta)-p'|$. A naive approach might take an average over the norms for all prediction-target pairs. However this gives rise to unwanted cross-terms and the possibility of multiple predictions matching the same target without any penalty for unmatched targets. Therefore, we choose a geometric mean over the predictions and an arithmetic mean over targets:
\begin{equation}
    \error(\theta, \targets) = \frac{1}{|\targets|}\sum_{p'\in\targets}
    \prod_{p(\theta)\in\predictions(\theta)}|p(\theta)-p'|
    ^{\frac{1}{|\predictions|}}
    \label{eq:error}
\end{equation}
The error term is only zero when each target is matched by at least one prediction and allows for cases where the number of predictions is greater than or equal to the number of targets $|\predictions| \geq |\targets|$. An alternative approach, which undesirably introduces more hyper-parameters, would be to let each prediction $\predictions(\theta)$ represent the centroid of a mixture distribution and use expectation-maximisation to match the centroids to targets $\targets$.

\subsubsection{Eigenvalue term: encouraging bifurcations}
\begin{wrapfigure}{r}{0.5\textwidth}
    \centering
    \includegraphics[width=\linewidth]{bifurcation-measure}
    \caption{Bifurcation measure $\measure(s)$ and determinant $\Det$ along the arclength $s$ of two different bifurcation curves demonstrating how maximising the measure along the curve maintains the existing bifurcation marked by a circle, while encouraging new bifurcations marked by stars.}
    \label{fig:measure}
\end{wrapfigure}

@@

@@

Equipped with a measure that quantifies the appearance of bifurcations along a bifurcation arc we can define the total measure for a bifurcation diagram as
\begin{equation}
    \Psi(\theta):=\frac{
        \int_{\rates(u,p)=0}\!
        \measure(s)\,\mathrm{d}s
    }{
        \int_{\rates(u,p)=0}\!\
        \mathrm{d}s
    }.
    \label{eq:total-measure}
\end{equation}
Here we denote $\int_{\rates(u,p)=0}\mathrm{d}s$ as the sum of the line integrals in $(u,p)\in\Reals^{N+1}$ defined by the level set $\rates(u,p)=0$ with $s$ being an arbitrary parametrisation of the curves. The total measure $\Psi(\theta)$ is normalised such that $\Psi(\theta)\rightarrow1$ in the regimes where the controlled condition region $p$ is densely packed with bifurcations. The total measure $\Psi(\theta)$ is added to the error term as if it were a likelihood. This defines the cost function as
\begin{align}
    \loss(\theta|\targets):=
    \big(|\predictions|-|\targets|\big)\log\Psi(\theta)+
    \error(\theta, \targets),
    \label{eq:loss}
\end{align}
The pre-factor $|\targets|-|\predictions|$ in the eigenvalue term ensures that the gradients are always pushing optimisers towards a state where $|\targets|=|\predictions|$. This can be seen as a step-wise annealing of the eigenvalue term until the desired state is reached.

\subsection{Differentiating the cost function}

To make use of gradient-based optimisers to locate desired bifurcation diagrams, we show here how to differentiate the cost function. First, we note that while individual bifurcations $p(\theta)$ depend smoothly on $\theta$, the total number of predictions $|\predictions|$ does not have gradient contributions with respect to $\theta$. Therefore, we can safely drop the dependency in the prediction counter and now proceed in taking gradients with respect to $\theta$ knowing that the only dependencies we need to track are for individual bifurcations $p(\theta)$ within the definition the error term \eqref{eq:error} and the total measure \eqref{eq:total-measure}. Therefore, 
\begin{align}
    \frac{\partial\loss}{\partial\theta}=
    \big(|\predictions|-|\targets|\big)\,\lambda\,
    \frac{\partial \Psi}{\partial\theta}\Psi(\theta)^{-1}+
    \frac{1}{|\targets||\predictions|}\sum_{p'}
    \prod_{p(\theta)}|p(\theta)-p'|^{\frac{1}{|\predictions|}}
    \sum_{p(\theta)}\frac{\partial p}{\partial\theta}\left(p(\theta)-p'\right)^{-1}
    \label{eq:gradient-loss}
\end{align}
In a similar vein to back-propagation through neural differential equations \cite{Chen2018NeuralEquations} we would like to be able to calculate the gradient $\frac{\partial\loss}{\partial\theta}$ without having to differentiate through the operations of the solver that finds the bifurcation diagram $\rates(u,p)=0$ and the bifurcation locations $p(\theta)$. To calculate the gradient of the measure $\frac{\partial \Psi}{\partial\theta}$ we need to differentiate line integrals that depend on $\theta$. Fortunately this can be done by the application of the generalised Leibniz integral rule, details of which can be found in Appendix \ref{appendix:leibniz-rule}.

@@

Our method can be used for more complex models than the minimal models shown in Figure \ref{fig:minimal-models:results}. To demonstrate this we applied the model to a 4 state 40 parameter model \cite{Grant2020InterpretationCircuit} (see Appendix \ref{appendix:double-exclusive}).

@@

We proposed a gradient-based approach for inferring the parameters of differential equations that produce a user-specified bifurcation diagram. By applying implicit layers \cite{Look2020DifferentiableLayers,Bai2019DeepModels} and the generalised Leibniz rule \cite{Flanders1973DifferentiationSign} to the geometry of the implicitly defined steady states \cite{Goldman2005CurvatureSurfaces} it is possible to use automatic differentiation methods to efficiently calculate gradients. We defined a bifurcation measure that uses the determinant of the state-space Jacobian as an indicator for bifurcating parameter regimes in the eigenvalue term of the cost function. The gradients of the cost can be efficiently computed using automatic differentiation methods. The computational bottleneck is the evaluation of the state-space Jacobian determinant which limits the implementation to ordinary differential equations.

@@

Although we did not consider limit cycles, the bifurcation measure can be extended to detect Poincar\'e-Andronov-Hopf bifurcations alongside changes in stability of fixed points (see Appendix \ref{appendix:hopf} for details).This measure enables detection of the onset of damped oscillations and/or the emergence of limit cycles (Figure \ref{fig:hopf-measure}). Used together with a steady state solver that detects periodic solutions and gradient-based optimisation, we can specify regions of damped oscillation and limit cycles. Our approach generalises naturally to bifurcation manifolds such as limit point curves or surfaces. This is because the normal components of implicit derivatives can still be calculated for under-determined systems of equations \cite{Jos2011OnSurface,Tao2016Near-IsometricTracking,Fujisawa2013CalculationInvariance}. In the case of manifolds it would be more appropriate to use isosurface extraction algorithms rather than continuation to obtain the steady-state manifold (further discussion in Appendix \ref{appendix:stedy-state-solvers}). Our approach does not depend on the details of the steady-state solver and therefore can still be applied.
